---
title: "Bibliometric part II"
author: "Kevin Lanning"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

## Some explorations of the nature and structure of computational social science using simple text analysis

As with the network visualizations, we use a tidy approach to text as well.

```{r setup, message=FALSE}
library(kableExtra) # for tables / requires install from github 2023
library(tidyverse)
library(tidytext)
dataDir <- "CompSocSci2023"
```

### first take: what are the top papers in each community (module) of the network?

```{r message=FALSE}
allNodes <- read_csv(file.path(dataDir,"CSSNodesText.csv"),
    col_types = cols()) %>% 
    select(Label = name,
           alltext, 
           TI,
           modularity_class = communityLouv, 
           page_rank=centralPR, 
           Degree=central0D)
# 
# get rid of unconnected papers 
mostNodes <- allNodes %>% 
    filter (Degree > 0) %>% 
    group_by(modularity_class) %>% 
    mutate(moduleSize = n()) %>% 
    ungroup() %>% 
# and communities with < 40 or so papers
    filter(moduleSize > (.05*nrow(allNodes)))

mostNodes %>%
    select(Label, TI, modularity_class, page_rank) %>%
    group_by(modularity_class) %>% 
    mutate(rank = rank(page_rank)) %>% 
    ungroup() %>% 
    filter(rank < 4) %>% 
    arrange(modularity_class, rank) %>% 
    kable (caption = "Three highest PR papers in top communities in core bibliometric network") %>%  
    kable_styling()
```

### Second take: what are the most common words in the whole dataset?

These use the tidytext approach to text analysis, which begins by restructuring texts as data frames. Both the size of this vector (the total number of words or tokens) and the first few terms are shown.

Then, we order the list of the top 100 terms, after dropping 'stop words' like "a" and "the".

```{r}
CSSwords <- allNodes %>% 
    select(alltext) %>% 
    unnest_tokens(word, alltext) 

# now the most common words in the whole dataset
wordList <- CSSwords %>%
    anti_join(get_stopwords()) %>% 
    count(word, sort = TRUE) 
wordList %>%
    slice(1:25) %>% 
    kable(caption = "Most common words in CSS papers") %>% 
          kable_styling()
```

### Third take: How does the language of different communities differ?

This identifies the characteristic language (in abstracts, titles, and keywords) of each of the larger communities identified via Gephi.

Stopwords are not dropped here, as these should wash in comparisons of different communities. Next, the long dataframe is reduced to a set of counts, then the word-counts are compared for the individual communites

```{r}
# the text of each paper is represented as a vector of words 
oneGrams <- mostNodes %>% 
    unnest_tokens(word, alltext) %>%
    filter(!grepl('[0-9]', word))
#    filter(is.na(as.numeric(word)))

# get overall counts of different words
oneGramAllCounts <-
    oneGrams %>% 
    count(word, sort = TRUE) %>%
    ungroup() 
head(oneGramAllCounts,3)

# get counts of words for each group
oneGramCommCounts <-
    oneGrams %>% 
    group_by(modularity_class) %>% 
    count(word, sort = TRUE) %>%
    ungroup() 

# combine the overall counts with the group counts
mostNodessumCount <- left_join(oneGramCommCounts,
                               oneGramAllCounts, 
                               by="word",
                               suffix = c("Comm",
                                          "Total")) %>%
# we drop words that appear only a few times in the data
# these are often names or neologisms found in just one source
            filter(nTotal > 3)
# this gets number of words (tokens) per group
mostNodesTotalWords <- mostNodessumCount %>% 
  group_by(modularity_class) %>% 
  mutate(total = sum(nComm))
# again a join to permit the tf/idf
CSStfIdf <- left_join(mostNodessumCount, 
          mostNodesTotalWords) %>% 
    bind_tf_idf(word,modularity_class,nComm) 
# and a listing of the top 25 terms in each
CSStfIdf %>%
  arrange(desc(tf_idf)) %>%
  group_by(modularity_class) %>% 
  top_n(25) %>% 
  select(word, modularity_class) %>% 
    aggregate(word ~ modularity_class,
              FUN = paste, 
              collapse = ", ") %>% 
    kable() %>% 
              kable_styling(position = "left", 
                  bootstrap_options = "striped",
                  font_size = 14)

```

## a few thoughts

We've used bibliometric data, network analysis, and text analysis to address the question of "what is computational social science?" How far have we gotten? Can, in principle, an empirical approach to the nature of "computational social science (CSS)" tell us what it is? How useful is bibliometric data? What other data sources might one use?

What are the biggest weaknesses you see in these analyses? How would you name the communities (or would you)?
