---
title: "From Modern, Appendix E"
output: html_document
date: "2026-03-17"
---

# correlation and regression
```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(mdsr)
library(mosaic)
glimpse(RailTrail)
```

## E.1.1 Motivating example: Modeling usage of a rail trail

```{r}
mod <- lm(volume ~ hightemp, data = RailTrail)
library(broom)
tidy(mod)
ggplot(RailTrail, aes(x = hightemp, y = volume)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

## E.1.2 Model visualization

```{r}
library(broom)
my_lm <- function(data, formula) { # native pipe fills in first unnamed argument
  lm(formula, data)
}
mod_avg <- RailTrail |>
  my_lm(volume ~ 1) |>
  augment(RailTrail)
mod_temp <- RailTrail |>
  my_lm(volume ~ hightemp) |>
  augment(RailTrail)
mod_data <- bind_rows(mod_avg, mod_temp) |>
  mutate(model = rep(c("null", "slr"), each = nrow(RailTrail)))
```

What is the point of this code? What do the two graphs describe?

```{r}
ggplot(data = mod_data, aes(x = hightemp, y = volume)) + 
  geom_smooth(
    data = filter(mod_data, model == "null"), 
    method = "lm", se = FALSE, formula = y ~ 1, 
    color = "dodgerblue", linewidth = 0.5
  ) + 
  geom_smooth(
    data = filter(mod_data, model == "slr"),
    method = "lm", se = FALSE, formula = y ~ x, 
    color = "dodgerblue", linewidth = 0.5
  ) + 
  geom_segment(
    aes(xend = hightemp, yend = .fitted), 
    arrow = arrow(length = unit(0.1, "cm")), 
    linewidth = 0.5, color = "darkgray"
  ) + 
  geom_point(color = "dodgerblue") + 
  facet_wrap(~model)
```

## E.1.3 Measuring the strength of fit

```{r}
n <- nrow(RailTrail)
SST <- var(pull(RailTrail, volume)) * (n - 1)
SSE <- var(residuals(mod)) * (n - 1)
1 - SSE / SST
```


## E.1.4 Categorical explanatory variables


What is the relationship between the values of the following two lines? Why? 

```{r}
coef(lm(volume ~ weekday, data = RailTrail))
RailTrail |>
  group_by(weekday) |>
  summarize(mean_volume = mean(volume))
```

E.2.1 Parallel slopes: Multiple regression with a categorical variable

From correlation to multiple regression (no interaction, parallel lines model)

```{r}
mod_parallel <- lm(volume ~ hightemp + weekday, data = RailTrail)
tidy(mod_parallel)
glance(mod_parallel)
```

```{r}
mod_parallel |>
  augment() |>
  ggplot(aes(x = hightemp, y = volume, color = weekday)) +
  geom_point() + 
  geom_line(aes(y = .fitted)) + 
  labs(color = "Is it a\nweekday?")
```

## E.2.3 Non-parallel slopes: Multiple regression with interaction

```{r}
mod_interact <- lm(volume ~ hightemp + weekday + hightemp * weekday, 
  data = RailTrail)
tidy(mod_interact)
```

## E.2.4 Modeling non-linear relationships

```{r}
mod_interact |>
  augment() |>
  ggplot(aes(x = hightemp, y = volume, color = weekday)) +
  geom_point() + 
  geom_line(aes(y = .fitted)) + 
  labs(color = "Is it a\nweekday?")
```

```{r}
ggplot(data = RailTrail, aes(x = hightemp, y = volume)) + 
  geom_point() +
  geom_smooth(method = lm) +
  geom_smooth(method = loess, color = "green") + 
  ylab("Number of trail crossings") + 
  xlab("High temperature (F)") 
```

10.2.1

```{r}
library(tidyverse)
library(mdsr)
url <- "http://archive.ics.uci.edu/static/public/2/adult.zip"
adult_zip <- tempfile()
download.file(url, destfile = adult_zip)
adult_dir <- here::here("data", "adult")
unzip(adult_zip, files = "adult.data", exdir = adult_dir)
adult_data <- here::here(adult_dir, "adult.data")

census <- adult_data |>
  read_csv(
    col_names = c(
      "age", "workclass", "fnlwgt", "education", 
      "education_1", "marital_status", "occupation", "relationship", 
      "race", "sex", "capital_gain", "capital_loss", "hours_per_week", 
      "native_country", "income"
    )
  ) |>
  mutate(income = factor(income))
glimpse(census)
```

## split into train and test

```{r}
library(tidymodels)
set.seed(364)
n <- nrow(census)
census_parts <- census |>
  initial_split(prop = 0.8)

train <- census_parts |>
  training()

test <- census_parts |>
  testing()

list(train, test) |>
  map_int(nrow)
```

## 10.2.1.1 the null model

simple percentage of high earners, then shown as table, then saved as a model

```{r}
(pi_bar <- train |>
  count(income) |>
  mutate(pct = n / sum(n)) |>
  filter(income == ">50K") |>
  pull(pct))

train |>
  count(income) |>
  mutate(pct = n / sum(n))

mod_null <- logistic_reg(mode = "classification") |>
  set_engine("glm") |>
  fit(income ~ 1, data = train)

library(yardstick)
pred <- train |>
  select(income, capital_gain) |>
  bind_cols(
    predict(mod_null, new_data = train, type = "class")
  ) |>
  rename(income_null = .pred_class)
accuracy(pred, income, income_null)

confusion_null <- pred |>
  conf_mat(truth = income, estimate = income_null)
confusion_null
```

## 10.2.1.2 Logistic regression 

```{r}
mod_log_1 <- logistic_reg(mode = "classification") |>
  set_engine("glm") |>
  fit(income ~ capital_gain, data = train)
train_plus <- train |>
  mutate(high_earner = as.integer(income == ">50K"))

ggplot(train_plus, aes(x = capital_gain, y = high_earner)) + 
  geom_count(
    position = position_jitter(width = 0, height = 0.05), 
    alpha = 0.5
  ) + 
  geom_smooth(
    method = "glm", method.args = list(family = "binomial"), 
    color = "dodgerblue", lty = 2, se = FALSE
  ) + 
  geom_hline(aes(yintercept = 0.5), linetype = 3) + 
  scale_x_log10(labels = scales::dollar)

```

