```{r, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, warn = FALSE, message = FALSE) # defaults for all chunks after this one
library(tidyverse)
library(HistData) # for Galton data
library(corrr) # for tidy corr matrix output
#library(dslabs)
#library(Ecfun)
#library(broom) # unlist regression results into tibbles
#library(modelr)
library(GGally) # for scatterplot matrix
library(Ecdat) # Econometric data incl. affairs dataset 
```

# from correlation to multiple regression

Please begin with a review of [Appendix E of the Modern text.](https://mdsr-book.github.io/mdsr3e/E-regression.html)

## the multivariate case

We begin with a second dataset. You can learn about it by typing `?swiss` in the console.

```{r swiss data}
data(swiss)
str(swiss)
```

Here's a *scatterplot matrix* of the Swiss data. Look at the first column of plots (or first row of the correlations). What is the relationship between fertility and each of the other variables?

```{r scatterplot matrix with ggpairs, swiss}
# ds_theme_set()
set.seed(0)
ggpairs (swiss,
        lower = list(
            continuous = "smooth"),
        axisLabels ="none",
        switch = 'both')
```

Here, we predict fertility from all of the remaining variables together in a single regression analysis, using the lm (linear model) command. Note that the result of this analysis is a list. We can pull out the key features of the data using the summary() command. How do you interpret this?

```{r multiple regression - swiss data}
swissReg <- lm(Fertility ~ ., data=swiss)
summary(swissReg)
```

Regression is a powerful tool for understanding the relationship between a response variable and one or more predictor variables. We can use it where our variables are not normally distributed, as in the case of dichotomous variables (yes/no, true/false), as well as counts, which often are skewed. Here's a second dataset, the marital affairs data, which is also included in the Ecdat package. We'll apply ggpairs here, but for clarity will show only half of the data at a time. The dependent variable of interest (nbaffairs) will be included in both plots:

```{r out.width = "100%", warning= FALSE, message = FALSE}
Fair1 <- Fair %>% 
  select(sex:child, nbaffairs)
ggpairs(Fair1,  
# if you wanted to jam all 9 vars onto one page you could do this
#        upper = list(continuous = wrap(ggally_cor, size = 10)),
         lower = list(continuous = 'smooth'),
         axisLabels = "none",
        switch = 'both')
Fair2 <- Fair %>% 
  select(religious:nbaffairs) 
ggpairs(Fair2, lower = list(continuous = 'smooth'),
           axisLabels = "none",
           switch = 'both')
```

```{r multiple regression - affairs}
affairReg <- lm(nbaffairs ~ ., data=Fair)
summary(affairReg)
```
