---
title: "firstgen_tidytext"
output: html_document
date: "2025-02-27"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidytext)
textsource <- read_csv("combinednonfirstgen.csv")
```

The code below uses the `tidytext` package to create a tidy text format of the data. This breaks up long strings of text into individual tokens (words) and creates a tidy data frame where each row corresponds to a single word from the original text. The `unnest_tokens` function is used for this purpose, which tokenizes the text in the `selftext` column of the data frame. The unnest_tokens function, by default, also removes punctuation and sets all words to lower case.

This tall dataframe is then reduced to a set of counts. These counts are saved to CSV files for further analysis.

```{r}
tidytext <- textsource %>% 
    unnest_tokens(word, selftext)
cleaned_text <- tidytext %>%
  anti_join(get_stopwords()) # for comparing the two corpora,
# you may or may not want to include this
counts <- cleaned_text %>%
  count(word, sort = TRUE)
counts |> write_csv("counts_nonfirstgen.csv")
cleaned_text |> write_csv("cleaned_nonfirstgen.csv")
tidytext |> write_csv("tidytext_nonfirstgen.csv")
```
